Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (*Sida fallax*)

### 2023-09-07 ms/ms.qmd
final checks before submission

```{=html}

<!-- TO DO,

1. at end of INTRODUCTION, address Tom's comments

"suggest you rigorously define sun leaves and shade leaves here, and dispense with the quotes."

Brief Communications—Brief Communications are short (c 3000–4000 word) research articles reporting exciting, significant new findings. They include no more than 4 figures and tables, combined. Manuscripts, and their abstracts, should be organized as described for Research articles. Authors must adhere to the same membership and page charge requirements as for regular Research Articles.

a. Title and Running Heads
The manuscript title for research papers should be specific and informative, conveying the key findings of the research in an active voice. Provide a title written with sentence-style capitalization. Below the title, list authors: each author’s first name, middle initial, surname. On the next line, give affiliation and unabbreviated address. If authors have different affiliations and addresses, add a superscript number after each author’s name to indicate the footnoted address. Include another footnote superscript number to indicate the author for correspondence.

Include the line: Manuscript received _______; revision accepted _______.

After this text, add a short title no longer than 65 characters (including spaces) to be used as a running head in the article.

b. Abstract
AJB requires structured abstracts for Research Articles and Brief Communications. The abstract is 250 words or fewer, written in the following structured format:
• Premise of the study (why the work was done, what major questions of plant biology are addressed, and why it is important to the broad AJB readership)
• Methods
• Key results
• Conclusions (what major points should the reader take from this article)

The abstract may be all a reader sees initially, so it should be written to capture the interest of the general botanical community as well as specialists. Avoid references; if essential, cite parenthetically with journal name, volume number, pages, and year.
-->
```
```{r, echo = FALSE, message = FALSE, warning = FALSE}

@@

@@

Amphistomy, the presence of stomata on both lower and upper surfaces of broad leaves, should increase carbon gain by reducing the average diffusion pathlength between stomata and chloroplasts (\autoref{fig:compare-circuit-diagrams}), yet paradoxically this seemingly simple adaptation is uncommon in nature and we don't know why. Understanding variation in stomatal traits like amphistomy is imperative because these tiny pores play an outsized ecological role in the global carbon and water cycles [@hetherington_role_2003; @berry_stomata:_2010]. A widely applicable, accurate representation of how stomata mediate the relationship between CO$_2$ gained through photosynthesis and water lost through evapotranspiration is essential to predict future climate using Earth Systems Models [@jarvis_interpretation_1976; @biggins_model_1987; @collatz_physiological_1991; @leuning_critical_1995; @sellers_modeling_1997]. Optimality models accurately predict the major cause of water loss, stomatal conductance ($g_\mathrm{sw}$), by assuming plants maximize carbon gain minus a cost of water [@cowan_stomatal_1977; @givnish_economy_1986; @medlyn_reconciling_2011; @lin_optimal_2015; @wang_towards_2017; @franks_comparing_2018; @deans_optimization_2020; @franklin_organizing_2020; @wang_theoretical_2020; @harrison_ecoevolutionary_2021]. Despite the success of optimality modeling in predicting $g_\mathrm{sw}$, the same modeling approach has so far failed to explain the rarity of amphistomatous leaves [@muir_is_2019]. **This gap between theory and observations strongly implies that we remain ignorant about some key benefits and costs associated with stomata.**

@@

The most promising adaptive hypothesis is that amphistomy is important for maximizing photosynthetic rate under high light. @mott_adaptive_1982 proposed that "plants with a high photosynthetic capacity, living in full-sun environments, and experiencing rapidly fluctuating or continuously available soil water" would benefit most, in terms of increased carbon gain, from having amphistomatous leaves. As described above, herbs from sunny habitats are often amphistomatous. Most variation in stomatal density ratio ($\mathrm{SR}$, the ratio of stomatal density between the upper and lower surfaces) among species is assumed to be genetic, but there is also putatively adaptive plasticity in response to light. Leaves of *Ambrosia cordifolia*, a desert perennial herb, are hypostomatous under low light (photosynthetic photon flux density, PPFD = 110 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) but develop \~20% of their stomata on the upper surface under high light (1700 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) [@mott_amphistomy_1991]. Similarly, *Solanum lycopersicum* leaves are hypostomatous when grown in the shade but develop \~20% of their stomata on the upper surface grown under high light-intensity [@gay_influence_1975]. Adult leaves of *Eucalyptus globulus* are amphistomatous, but the proportion of adaxial stomata increases from \~10-20% under low light to \~30-40% under high light [@james_leaf_2001]. In summary, both genetic and plastic responses to evince a widespread association between light and $\mathrm{SR}$.

The association between high light and amphistomy suggests that 'sun' leaves have the most to gain in terms of increased photosynthesis from having stomata on both surfaces, as @mott_adaptive_1982 hypothesized. @parkhurst_adaptive_1978 proposed quantifying this benefit as 'amphistomy advantage' ($\mathrm{AA}$), which we adopt here with some modification (see Materials and Methods). This hypothesis has never been tested directly by comparing the photosynthetic rate of an amphistomatous leaf to that of an otherwise identical hypostomatous leaf with the same total stomatal conductance under the same conditions. We propose a straightforward method to do this by experimentally creating a pseudohypostomatous leaf with gas exchange blocked through the upper surface (see Materials and Methods). One reason that sun leaves might have greater $\mathrm{AA}$ is that they are usually thicker or denser [@poorter_metaanalysis_2019], resulting in lower $g_\mathrm{ias}$ [@parkhurst_adaptive_1978]. A nonmutually exclusive hypothesis is that if sun leaves have a stomatal density ratio closer to 0.5 (same density on each leaf surface), this will confer a greater advantage than an amphistomatous leaf with most stomata on one surface. In other words, amphistomy doesn't make much difference if one leaf surface has very open stomata on it. We therefore predict that sun leaves will have greater $\mathrm{AA}$ possibly because they have thicker leaves and/or $\mathrm{SR}$ closer to 0.5. We actually report $g_{\mathrm{smax,ratio}}$, which is similar to $\mathrm{SR}$ except that takes into account differences in both stomatal density and size between surfaces.

The native flora of the Hawaiian archipelago is a excellent system to test the relationship between light habitat and $\mathrm{AA}$. Many lineages have adapted to different light habitats after colonization and leaf anatomical traits such as $\mathrm{SR}$ and thickness vary within and among closely related species. It is hypothesized that the common ancestor in many Hawaiian clades was a "weedy" species with high dispersal ability adapted to open habitats [@carlquist_biota_1966]. Colonization was followed by adaptive radiation into higher elevation, montane, closed, forested habitats. Consequently, adaptation to sun and shade is a common axis of phenotypic variation among Hawaiian plants such as lobeliads [@givnish_adaptive_2004; @montgomery_adaptive_2008; @givnish_origin_2009; @givnish_common-garden_2014; @scoffoni_light-induced_2015], *Bidens* [@carlquist_biota_1966; @knope_dispersal_2020], *Scaevola* [@robichaux_evolution_1984; @mckown_trait_2016], *Euphorbia* [@sporck_hawaiian_2011], and *Plantago* [@dunbar-co_leaf_2009].

Here we focus on variation within an indigenous plant species *Sida fallax* Walp. (Malvaceae), known in the Hawaiian language as ʻilima. ʻIlima is found from sea level to elevations $>1000$ mas on multiple Hawaiian islands. Coastal populations are morphologically different from montane populations (\autoref{fig:study-system}). Coastal regions of Hawaiʻi are characterized by high sun exposure, warmer temperatures, high winds, salinity, and variation in water availability. Coastal populations of ʻilima tend to be short and prostrate which likely helps them to withstand the windy environment (\autoref{fig:study-system}a). The leaves of these populations are covered on both surfaces in dense, soft hairs that give the leaves a silvery green appearance (\autoref{fig:study-system}b), which helps mitigate water loss by reflecting solar radiation, thereby lowering leaf temperature [@ehleringer_pubescence_1978]. Montane regions, on the other hand, provide very different challenges. Many other tall species grow on the slopes of these wet mountainous regions, which makes light competition a factor that plants may need to adapt to. Possibly due to this, montane populations are erect and shrub- or tree-like, capable of growing meters tall with strong, woody stems. These individuals have smooth, green foliage with serrated edges. Montane populations exhibit traits that help them to compete for light availability (Raunch et al. 1997). This montane morphology is not found in *S. fallax* populations on other Pacific Islands [@pejhanmehr_phylogenetic_2022]. The variable morphology between coastal and montane populations has even led some scientists to argue that it makes up a species complex, rather than a single species (Raunch et al. 1997).

The objectives of our study are to test whether 1) 'sun' leaves of coastal ʻilima plants will have greater $\mathrm{AA}$ than 'shade' leaves of montane plants; and if so, is this because 2a) coastal plants have thicker leaves than montane plants and/or 2b) coastal plants have a $g_{\mathrm{smax,ratio}}$ closer to 0.5?

```{=html}
<!-- TEXT FROM PROPOSAL THAT MAY HAVE SOMETHING USEFUL:

*Knowing the adaptive significance of amphistomy will advance our understanding of leaf structure-function relationships, aid reconstructing paleoenvironments, and guide crop optimization*. Stomata are studied in disparate fields such as cell biology, ecosystem ecology, and paleontology [@hetherington_role_2003; @bergmann_stomatal_2007; @berry_stomata:_2010; @mcelwain_paleoecology_2017]. The intense focus on stomata makes it all the more surprising that we know relatively little about the adaptive significance of where stomata are on the leaf! Amphistomy is deeply integrated with leaf traits associated with high photosynthetic capacity and leaf water transport capacity [@smith_leaf_1997; @drake_two_2019; @xiong_one_2020]. Knowing the upper limits of photosynthesis will help breed more productive crops [@long_can_2006] and predict future terrestrial carbon sinks at higher atmospheric CO$_2$ [@keenan_terrestrial_2018]. In paleontology, stomata visible on fossil leaves help reconstruct paleo-CO$_2$ concentrations based on density/index [@konrad_modelling_2008; @franks_co2-forced_2009; @mcelwain_paleoecology_2017; but see @jordan_links_2020], but stomatal density ratio ($\mathrm{SR}$) may add information about whether plants lived in open or closed vegetation [@salisbury_i_1928; @jordan_using_2014; @bucher_stomatal_2017]. Finally, $\mathrm{SR}$ tends to increase (higher proportion of stomata on upper surface) during domestication [@milla_shifts_2013], but we do not know why. Our work will help infer how changes in $\mathrm{SR}$ and other leaf anatomical traits affect photosynthesis and water use in crops. This is particularly promising because we may learn ways to increase leaf photosynthesis without increasing water loss by altering $\mathrm{SR}$ and/or the regulation of stomatal aperture on each leaf surface.

To redress these gaps in our knowledge, in Aim 1 we will model the benefits and costs of amphistomy under low- and high-light conditions and in Aims 2-3 we will test model predictions using the wild tomatoes as model clade (*Solanum* spp.; Solanaceae). Specifically, we will experimentally manipulate whether gas exchange proceeds through one or both leaf surfaces, holding all else constant. This will enable us to measure how much amphistomy increases photosynthesis compared to an otherwise identical hypostomatous leaf (Aim 2). In Aim 3, we will test whether amphistomy is costly in shade leaves because it causes dehydration of the palisade mesophyll under high evaporative demand. The result of this project will determine whether high light, photosynthetic capacity, and adequate hydration are sufficient to explain the distribution of amphistomatous leaves in nature, as Mott and colleagues hypothesized 40 years ago.

The dearth of amphistomatous leaves implies a fitness cost associated with stomata on the upper leaf surface in particular [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984-1; @muir_is_2019]. Two nonmutually exclusive hypotheses for the cost of upper stomata have received the most attention, but neither has been tested experimentally. We will test one hypothesis, which is that amphistomy hinders photosynthesis by causing dehydration of the palisade mesophyll under high evaporative demand [@buckley_sites_2017; @drake_two_2019]. This cost of amphistomy could be greater in shade leaves because they dehydrate severely when exposed to intense sunflecks, warm temperatures, and/or low humidity, whereas sun leaves have constitutive traits that mitigate the risk of dehydration. Sensitivity to dehydration would be more consequential if it affected the palisade mesophyll, which is adjacent to the illuminated upper surface and is thus more photosynthetically active [@smith_leaf_1997; @evans_profiles_2003; @earles_embracing_2019] than the spongy mesophyll, which is adjacent to the lower surface (\autoref{fig:compare-circuit-diagrams}). An alternative hypothesis with some empirical support is that stomata on the upper surface are more likely to be colonized by bacterial or fungal pathogens, presumably because a propagule is more likely to land on the upper surface [@gutschick_photosynthesis_1984-1; @mckown_association_2014; @mckown_role_2019; @fetter_growthdefense_2021]. The current proposal does not address this pathogen susceptibility hypothesis because PI Muir is seeking funding for this work elsewhere. Although not the focus of our aims, our modeling and data can address two less well theorized alternative hypotheses that upper stomata compete for space on the epidermis [@de_boer_optimal_2016; @haworth_allocation_2018; @muir_how_2023] or alter palisade mesophyll structure [@dow_disruption_2017; @lundgren_mesophyll_2019; @baillie_developmental_2020].

*Despite the fact that we do not know the adaptive significance of amphistomy, the distribution of hypo- and amphistomatous leaves is nonrandom and most likely caused by natural selection.* Consistent relationships between stomatal traits, growth form, and light environment strongly suggest convergent evolution, a hallmark of adaptive evolution. Hypo- and amphistomatous leaves are two ends of a continuum in $\mathrm{SR}$. Hypostomatous leaves have a $\mathrm{SR}$ close to 0; fully amphistomatous leaves with equal densities on each leaf have a $\mathrm{SR}$ close to 0.5. Herbaceous plants from high-light habitats tend to be amphistomatous and have higher $\mathrm{SR}$ values than herbaceous plants from shaded habitats or woody species [@salisbury_i_1928; @parkhurst_adaptive_1978; @mott_adaptive_1982; @parkhurst_intercellular_1990; @peat_comparative_1994; @muir_making_2015; @bucher_stomatal_2017; @muir_light_2018]. A second line of evidence for adaptation is that the distribution of $\mathrm{SR}$ among species is inconsistent with the expected distribution for a ratio trait evolving neutrally; instead, a nonneutral bimodal distribution evolves repeatedly in land plants [@muir_making_2015]. Finally, amphistomatous leaves are strongly coordinated with other traits involved with high photosynthetic capacity [@beerling_evolutionary_1996; @xiong_one_2020] and domestication [@milla_shifts_2013], suggesting that amphistomy is favored by selection in some circumstances. What specific fitness costs and benefits underlie these adaptive patterns?

*The most promising adaptive hypothesis is that amphistomy is important for maximizing photosynthetic rate under high light*. To reiterate, herbaceous leaves in high light habitats tend to be amphistomatous. Most variation in $\mathrm{SR}$ among species is assumed to be genetic, but there is also putatively adaptive plasticity in response to light. Leaves of *Ambrosia cordifolia*, a desert perennial herb, are hypostomatous under low light (photosynthetic photon flux density, PPFD = 110 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) but develop ~20% of their stomata on the upper surface under high light (1700 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) [@mott_amphistomy_1991]. Similarly, *Solanum lycopersicum* leaves are hypostomatous when grown in the shade but develop ~20% of their stomata on the upper surface grown under high light-intensity [@gay_influence_1975]. Adult leaves of *Eucalyptus globulus* are amphistomatous, but the proportion of upper stomata increases from ~10-20% under low light to ~30-40% under high light [@james_leaf_2001]. In summary, both genetic and plastic responses to evince a widespread association between light and $\mathrm{SR}$.

*We don’t know whether amphistomy is an adaptation to high light intensity because no one has shown experimentally that amphistomy* per se *increases CO$_2$ assimilation more under high light than under low light*. Other factors correlated with light habitat could drive selection on $\mathrm{SR}$ and result in a spurious association between light habitat and $\mathrm{SR}$. Even if light intensity is the primary causal selective agent for $\mathrm{SR}$, its effect on selection for higher/lower $\mathrm{SR}$ could be the result of 1) *facultative* responses to light intensity and/or 2) *constitutive* differences between shade and sun leaves (\autoref{fig:causal-graph}). The *facultative* hypothesis is that amphistomy increases photosynthesis under high light intensity because of high CO$_2$ demand and/or CO$_2$ supply is constrained by low $g_\text{ias}$. Under this hypothesis, the advantage of amphistomy depends on the immediate light intensity, not the light environment in which plants develop. The nonmutually exclusive *constitutive* hypothesis is that developmental plasticity results in 'sun' and 'shade' leaves with manifold differences in anatomy and biochemistry such as thicker leaves, denser cell packing, or increased Rubisco concentration [@terashima_why_2001; @lambers_plant_2008]. Sun leaves benefit more from amphistomy than shade leaves under all light intensities because they have lower $g_\text{ias}$ and/or greater CO$_2$ demand. The proposed research will quantify the relative contributions of facultative and constitutive effects on the photosynthetic advantage of amphistomy. We will describe predictions of these hypotheses in more detail below.

MOVE THIS TO INTRO Parkhurst (1978) introduced the concept of “amphistomy advantage” to quantify the photosynthetic advantage of an amphistomatous leaf over an otherwise identical hypostomatous leaf. An important limitation of his approach is that he assumed photosynthesis was simply proportional to the total diffusive conductance from the atmosphere to the sites of carboxylation; that tacitly assumes CO$_2$ concentration at the sites of carboxylation is zero, which in turn infinite local did not model assumes CO$_2$ demand is locally infinite throughout the leaf. In reality, photosynthesis at any given point in the leaf depends on the local concentrations of CO$_2$, chlorophyll, and Rubisco, and the local irradiance. Thus, some mesophyll layers can draw down chloroplastic CO$_2$ concentration to a low level and potentially assimilate a lot of carbon when CO$_2$ is adequately supplied, whereas other layers have much weaker CO$_2$ demand. By not incorporating a CO$_2$ demand function, Parkhurst assumed that any increase in CO$_2$ supply was matched by a concomitant increase in CO$_2$ assimilation. We now know that increasing assimilation depends on both CO$_2$ supply and demand. If supply increases, but the demand function is flat because of low light intensity and/or low photosynthetic capacity, then $A$ won’t change much (demand-limited). Conversely, if the demand function is steep because there is excess light and/or enzymatic capacity, then increased CO$_2$ supply will yield a large increase in $A$ (supply-limited).

 -->
```
```{=tex}
\begin{figure}[H]
  \includegraphics{../figures/study-system.pdf}
  \caption{(A) Typical growth form of montane (left) and coastal (right) ‘ilima plants and (B) leaves. (C) Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).}
  \label{fig:study-system}
\end{figure}
```
# MATERIALS AND METHODS --- {#methods}

@@

We identified 8 suitable natural populations of ʻilima on Oʻahu and 5 on Hawaiʻi Island by consulting [@yorkston_interfertility_2006] and citizen scientist records on iNaturalist [@noauthor_inaturalist_2022] (\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites that appeared to be cultivated. We visited sites between August and November 2022. For logistical reasons, the sites on Hawaiʻi were sampled in one three-day trip. We haphazardly sampled eight plants distributed evenly between the highest and lowest elevation plants along a transect at each site. For safety and conservation reasons, transects were along a trail or road. We did not sample small individuals if there was risk removing leaves would cause mortality. From each plant, we collected two fully expanded leaves for traits measurements. We sampled stomatal traits on all leaves; leaf thickness on one leaf from three randomly selected plants per site; and, due to limited time, a single leaf from a single plant at the middle of each transect for gas exchange measurements.

@@

We cut thin sections using two razor blades taped together and cutting the leaf in a petri dish of water, wet-mounted sections onto a slide, and took digital micrographs using a brightfield microscope, as described above.

@@

At each site, we selected one representative leaf from one plant near the middle of the transect for gas exchange measurements using a portable infrared gas analyze LI-6800PF, LI-COR Biosciences, Lincoln, NE, USA). We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at saturating light ($\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), ambient CO$_2$ (415 ppm), and $T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}$. We also estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modification. For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values, holding other environmental variables constant. We refer to this as an $A \textendash g_\text{sw}$ curve. We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of "pseudohypostomatous" leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm). We refer to the untreated and partially blocked leaves as "amphi" and "pseudohypo", respectively. $\mathrm{AA}$ is calculated as the log-response ratio of $A$ compared at the same total $g_\text{sw}$:

@@

To control for order effects, we alternated between starting with amphi (untreated) or pseudohypo (treated) leaf measurements. In the field, we acclimated the focal leaf to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), high relative humidity ($\mathrm{RH} = 70\%$), $T_\text{leaf} = 25-30~° \text{C}$, depending on ambient temperature, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without photosynthetic downregulation. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process with the other treatment (amphi or pseudohypo).

@@

We tested whether leaf thickness (log-transformed) varied between coastal and montane populations and among individuals within populations using a Bayesian mixed-effects model with site-type as a fixed effect and individual plant and site as random effects. We used the *R* package **brms** version `r fit_thickness$version$brms` [@burkner_brms_2017] to fit the model in *Stan* [@stan_development_team_stan_2023] with CmdStan version `r fit_thickness$version$cmdstan` and **cmdstanr** version `r fit_thickness$version$cmdstanr` [@gabry_cmdstanr_2023]. We sampled the posterior distribution from `r dim_fit_thickness[2]` chains with `r dim_fit_thickness[1]` iterations each after `r dim_fit_thickness[1]` warmup iterations. We estimated the relationship between population average leaf thickness and $\mathrm{AA}$ measured from a single individual per population. We used this approach most of the variation in leaf thickness was among sites (see Results) and the plant selected for gas exchange measurements was not always among the plants randomly selected for leaf thickness, precluding individual level correlation. We propagated uncertainty about in $\mathrm{AA}$ and leaf thickness by integrating over the entire posterior distribution sample for each variable. The key prediction is that the effect of leaf thickness on $\mathrm{AA}$ is positive.

@@

Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo with a DOI and stable URL upon publication. Raw data will be deposited on Dryad with a DOI and stable URL upon publication.

@@

Additional supporting information may be found online in the Supporting Information section at the end of the article

>>>>

```{r, echo = FALSE, message = FALSE, warning = FALSE}

@@

AA_coastal1 = pretty_est(habitat_aa1$coastal,
                        habitat_aa1$coastal.lower,
                        habitat_aa1$coastal.upper,
                        digits = 2L)

@@

Amphistomy, the presence of stomata on both lower and upper surfaces of broad leaves, should increase carbon gain by reducing the average diffusion pathlength between stomata and chloroplasts, yet paradoxically this seemingly simple adaptation is uncommon in nature and we don't know why. Understanding variation in stomatal traits like amphistomy is imperative because these tiny pores play an outsized ecological role in the global carbon and water cycles [@hetherington_role_2003; @berry_stomata:_2010]. A widely applicable, accurate representation of how stomata mediate the relationship between CO$_2$ gained through photosynthesis and water lost through evapotranspiration is essential to predict future climate using Earth Systems Models [@jarvis_interpretation_1976; @biggins_model_1987; @collatz_physiological_1991; @leuning_critical_1995; @sellers_modeling_1997]. Optimality models accurately predict the major cause of water loss, stomatal conductance ($g_\mathrm{sw}$), by assuming plants maximize carbon gain minus a cost of water [@cowan_stomatal_1977; @givnish_economy_1986; @medlyn_reconciling_2011; @lin_optimal_2015; @wang_towards_2017; @franks_comparing_2018; @deans_optimization_2020; @franklin_organizing_2020; @wang_theoretical_2020; @harrison_ecoevolutionary_2021]. Despite the success of optimality modeling in predicting $g_\mathrm{sw}$, the same modeling approach has so far failed to explain the rarity of amphistomatous leaves [@muir_is_2019]. **This gap between theory and observations strongly implies that we remain ignorant about some key benefits and costs associated with stomata.**

@@

The most promising adaptive hypothesis is that amphistomy is important for maximizing photosynthetic rate under high light. @mott_adaptive_1982 proposed that "plants with a high photosynthetic capacity, living in full-sun environments, and experiencing rapidly fluctuating or continuously available soil water" would benefit most, in terms of increased carbon gain, from having amphistomatous leaves. As described above, herbs from sunny habitats are often amphistomatous. Most variation in stomatal density ratio ($\mathrm{SR}$, the ratio of stomatal density between the upper and lower surfaces) among species is assumed to be genetic, but there is also putatively adaptive plasticity in response to light. Leaves of *Ambrosia cordifolia*, a desert perennial herb, are hypostomatous under low light (photosynthetic photon flux density, PPFD = 110 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) but develop \~20% of their stomata on the upper surface under high light (1700 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$) [@mott_amphistomy_1991]. Similarly, *Solanum lycopersicum* leaves are hypostomatous when grown in the shade but develop \~20% of their stomata on the upper surface grown under high light-intensity [@gay_influence_1975]. Adult leaves of *Eucalyptus globulus* are amphistomatous, but the proportion of adaxial stomata increases from \~10-20% under low light to \~30-40% under high light [@james_leaf_2001]. In summary, both genetic and plastic responses evince a widespread association between light and $\mathrm{SR}$.

The association between high light and amphistomy suggests that 'sun' leaves have the most to gain in terms of increased photosynthesis from having stomata on both surfaces, as @mott_adaptive_1982 hypothesized. @parkhurst_adaptive_1978 proposed quantifying this benefit as 'amphistomy advantage' ($\mathrm{AA}$), which we adopt here with some modification (see Materials and Methods). This hypothesis has never been tested directly by comparing the photosynthetic rate of an amphistomatous leaf to that of an otherwise identical hypostomatous leaf with the same total stomatal conductance under the same conditions. We propose a straightforward method to do this by experimentally creating a pseudohypostomatous leaf with gas exchange blocked through the upper surface (see Materials and Methods). One reason that sun leaves might have greater $\mathrm{AA}$ is that they are usually thicker or denser [@poorter_metaanalysis_2019], resulting in lower $g_\mathrm{ias}$ [@parkhurst_adaptive_1978]. A nonmutually exclusive hypothesis is that if sun leaves have a stomatal density ratio closer to 0.5 (same density on each leaf surface), this will confer a greater advantage than an amphistomatous leaf with most stomata on one surface. In other words, amphistomy doesn't make much difference if one leaf surface has few open stomata on it. We therefore predict that sun leaves will have greater $\mathrm{AA}$ possibly because they have thicker leaves and/or $\mathrm{SR}$ closer to 0.5. We actually report $g_{\mathrm{smax,ratio}}$, which is similar to $\mathrm{SR}$ except that it accounts for differences in both stomatal density and size between surfaces.

The native flora of the Hawaiian archipelago is a excellent system to test the relationship between light habitat and $\mathrm{AA}$. Many lineages have adapted to different light habitats after colonization and leaf anatomical traits such as $\mathrm{SR}$ and thickness vary within and among closely related species. It is hypothesized that the common ancestor in many Hawaiian clades was a weedy species with high dispersal ability adapted to open habitats [@carlquist_biota_1966]. Colonization was followed by adaptive radiation into higher elevation, montane, closed, forested habitats. Consequently, adaptation to sun and shade is a common axis of phenotypic variation among Hawaiian plants such as lobeliads [@givnish_adaptive_2004; @montgomery_adaptive_2008; @givnish_origin_2009; @givnish_common-garden_2014; @scoffoni_light-induced_2015], *Bidens* [@carlquist_biota_1966; @knope_dispersal_2020], *Scaevola* [@robichaux_evolution_1984; @mckown_trait_2016], *Euphorbia* [@sporck_hawaiian_2011], and *Plantago* [@dunbar-co_leaf_2009].

Here we focus on variation within an indigenous plant species *Sida fallax* Walp. (Malvaceae), known in the Hawaiian language as ʻilima. ʻIlima is found from sea level to elevations $>1000$ mas on multiple Hawaiian islands. Coastal populations are morphologically different from montane populations (\autoref{fig:study-system}). Coastal regions of Hawaiʻi are characterized by high sun exposure, warmer temperatures, high winds, salinity, and variation in water availability. Coastal populations of ʻilima tend to be short and prostrate which likely helps them to withstand the windy environment (\autoref{fig:study-system}a). The leaves of these populations are covered on both surfaces in dense, soft hairs that give the leaves a silvery green appearance (\autoref{fig:study-system}b), which helps mitigate water loss by reflecting solar radiation, thereby lowering leaf temperature [@ehleringer_pubescence_1978]. Montane regions, on the other hand, provide very different challenges. Many other tall species grow on the slopes of these wet mountainous regions, which makes light competition a factor that plants may need to adapt to. Possibly due to this, montane populations are erect and shrub- or tree-like, capable of growing meters tall with strong, woody stems. These individuals have smooth, green foliage with serrated edges. Montane populations exhibit traits that may help them to compete for light availability. This montane morphology is not found in *S. fallax* populations on other Pacific Islands [@pejhanmehr_phylogenetic_2022].

Because of their contrasting habitat and morphology, we treat leaves from coastal and montane plants as representatives of sun and shade leaves, respectively, for testing hypotheses about amphistomy advantage. Specifically, the objectives of our study are to test whether 1) sun leaves of coastal ʻilima plants will have greater $\mathrm{AA}$ than shade leaves of montane plants; and if so, is this because 2a) coastal plants have thicker leaves than montane plants and/or 2b) coastal plants have a $g_{\mathrm{smax,ratio}}$ closer to 0.5?

```{=tex}
\begin{figure}[H]
  \includegraphics{../figures/study-system.pdf}
  \caption{(A) Typical growth form of montane (left) and coastal (right) ‘ilima plants and (B) leaves. (C) Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).}
  \label{fig:study-system}
\end{figure}
```

# MATERIALS AND METHODS --- {#methods}

@@

We identified 7 suitable natural populations of ʻilima on Oʻahu and 5 on Hawaiʻi Island by consulting @yorkston_interfertility_2006 and citizen scientist records on iNaturalist [@noauthor_inaturalist_2022] (\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites that appeared to be cultivated. We visited sites between August and November 2022. For logistical reasons, the sites on Hawaiʻi were sampled during one three-day trip. We haphazardly sampled eight plants distributed evenly between the highest and lowest elevation plants along a transect at each site. For safety and conservation reasons, transects were along a trail or road. We did not sample small individuals if there was risk removing leaves would cause mortality. From each plant, we collected two fully expanded leaves for trait measurements. We sampled stomatal traits on all leaves; leaf thickness on one leaf from three randomly selected plants per site; and, due to limited time, a single leaf from a single plant at the middle of each transect for gas exchange measurements.

@@

We cut thin sections using two razor blades taped together. We sectioned the leaf in a petri dish of water, wet-mounted sections onto a slide, and took digital micrographs using a brightfield microscope, as described above.

@@

At each site, we selected one representative leaf from one plant near the middle of the transect for gas exchange measurements using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, NE, USA). We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at saturating light ($\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), ambient CO$_2$ (415 ppm), and $T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}$. We also estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modification. For each leaf, we measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values. We refer to this as an $A \textendash g_\text{sw}$ curve. We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm). Hypostomy refers to leaves with stomata only present on the lower, typically abaxial, surface. We refer to the untreated and partially blocked leaves as "amphi" and "pseudohypo", respectively. $\mathrm{AA}$ is calculated as the log-response ratio of $A$ compared at the same total $g_\text{sw}$:

@@

To control for order effects, we alternated between starting with amphi (untreated) or pseudohypo (treated) leaf measurements. In the field, we acclimated the focal leaf to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), high relative humidity ($\mathrm{RH} = 70\%$), $T_\text{leaf} = 25-30~° \text{C}$, depending on ambient temperature, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without photosynthetic downregulation. All other environmental conditions in the leaf chamber remained the same. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process with the other treatment (amphi or pseudohypo).

@@

We tested whether leaf thickness (log-transformed) varied between coastal and montane populations and among individuals within populations using a Bayesian mixed-effects model with habitat as a fixed effect and individual plant and site as random effects. We used the *R* package **brms** version `r fit_thickness$version$brms` [@burkner_brms_2017] to fit the model in *Stan* [@stan_development_team_stan_2023] with CmdStan version `r fit_thickness$version$cmdstan` and **cmdstanr** version `r fit_thickness$version$cmdstanr` [@gabry_cmdstanr_2023]. We sampled the posterior distribution from `r dim_fit_thickness[2]` chains with `r dim_fit_thickness[1]` iterations each after `r dim_fit_thickness[1]` warmup iterations. We estimated the relationship between population average leaf thickness and $\mathrm{AA}$ measured from a single individual per population. We used this approach most of the variation in leaf thickness was among sites and the plant selected for gas exchange measurements was not always among the plants randomly selected for leaf thickness, precluding individual level correlation. We propagated uncertainty about in $\mathrm{AA}$ and leaf thickness by integrating over the entire posterior distribution sample for each variable. The key prediction is that the effect of leaf thickness on $\mathrm{AA}$ is positive.

@@

Custom scripts are available on a GitHub repository (https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo with a DOI and stable URL upon publication. Raw data will be deposited on Dryad with a DOI and stable URL upon publication. [THE GITHUB REPO AND DRYAD DATA ARE AVAILABLE TO REVIEWERS]

@@

Additional supporting information may be found online in the Supporting Information section at the end of the article.
